What We Eval in the Shadows

### 2021-09-13 paper/main.tex
Fix an ugle page break interfering with a code block
Fix issues in 6

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute}, and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected the top 117 sites, which contribute to 90\% of the calls in
the corpus. We present examples from nine classes from this batch. The
classification is not exhaustive as we only study a fraction from the
\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit
in a specific category. Instead, the goal is to demonstrate the wide variety of
uses to which \eval is put to use. At the high level, the usage can be grouped
into two groups: one contains patterns that aim to provide better
API and the other focuses on simplifying implementation.

\subsection{Better API}

This category contains \eval used to develop a better API in terms of
ergonomics, extensibility, and ease of use. This is important given
that R is often being used interactively for data exploration.

@@

\begin{lstlisting}[title={\k{data.table::[}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
#    origin       V1
# 1:    LGA 11.35083
# 2:    EWR 14.96171
# 3:    JFK 12.45414
\end{lstlisting}\medskip

@@

\enlargethispage{2mm}
\begin{lstlisting}[title={\k{glue::glue_data}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
# JFK has average delay of 12 min
\end{lstlisting}

@@

\paragraph{Extracting Varargs} \Eval is used to obtain the source code of the
arguments passed to a vararg parameter (\k{...}). Concretely,
\k{substitute(alist(...))} yields an expression representing a call to
\k{alist} with the \k{...} expanded to the arguments supplied to the function.
Passing this to \eval returns a list of the unevaluated argument
expressions. In the following example, it is used to find and return the value of the first expression that evaluates to non-null value. The extracting varargs pattern is needed to avoid premature promise forcing.

\begin{lstlisting}[title={\k{statnet.common::NVL}}, captionpos=b]
 function(...) {
   for (e in eval(substitute(alist(...)))) {
     if (!is.null(eval(e, parent.frame()))) break
   }
   e
 }
\end{lstlisting}

\paragraph{Controlled Evaluation} \Eval can be used to control when and if
expressions are evaluated. For example, \k{assertthat} provides users with the
ability to make assertions and, if they fail, produce legible error messages.
The \texttt{\small assertthat::see\_if} function accepts a vararg with a list of
expressions. It extracts these expressions using the extracting varargs
pattern and evaluates them sequentially in the caller's environment using
\k{eval}. If an expression does not hold, an error is generated.

@@

\paragraph{Shadowing} Some packages use \eval to implement variations of the
\k{base} R functions. For example, the \k{igraph} package provides the
\k{do_call} function shown below. This is a variant of the built-in \k{do.call}
function with different semantics for resolving the function name. It
constructs a call from a name and a list of arguments and passes it to \eval.

@@

The patterns in this category use \eval to simplify implementation in terms of
avoiding repetitive and boilerplate code.

\paragraph{Code Generation} \Eval can be used to generate code that would
either be tedious to write by hand or that is only known after deployment. The
following code snippet for the \k{Rcpp} package uses \eval to automatically
generate R bindings to C++ methods.

@@

\paragraph{Boilerplate Removal} A variant of code generation is to let users
write compact code that is expended via \eval and \k{match.call}. The example
uses \k{match.call} to access the expression with which the function is called.
The following three lines construct a call to \k{stats::model.frame} with
argument text of parameters \k{'formula'}, \k{'data'}, and \k{'weights'}. Then,
the call expression is evaluated in the parent environment. The use of
\k{match.call} avoids repeated use of \k{substitute} for extracting the
argument expressions. It is extensible as new parameter names can be easily
added. This example illustrates one of the most recurring uses of \eval by
packages that do statistical modeling, and relates to the \k{model.frame} shape
in Table~\ref{tab:minimizedexpressions}.

@@

\paragraph{Obfuscation}  A few packages use \eval to bypass the restrictions
imposed by the \k{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. Package authors use \eval to obfuscate their code to get around this
limitation. The example below mutates the variable \k{s} in the package
environment. However, that is locked by default and unlocking it requires
\k{unlockBinding} which is a restricted function. To work around this, the call
to \k{unlockBinding} is done through \eval.

@@

In R, \eval is used chiefly for meta-programming and accessing remote
environments. Unlike JavaScript, which according to~\citet{ecoop11} had a small
set of well-defined patterns that could be rewritten without \eval, our results
suggest that \eval is an integral part of programming in R. However, there are
cases in which \eval is not needed, and other, more stylized functions could
have been used. There are also alternatives to \eval that provide
a more principled way for meta-programming and dealing with lazy evaluation. In
this section, we briefly discuss these issues.

\paragraph{Unnecessary use of eval} The cases where \eval appears to be an
overkill include simple expression shapes such as \k{X}, \k{V}, \texttt{\$}, \k{<-},
and \k{FUN}. Variable lookup in a remote environment can be performed by the
built-in \k{get} function. A similar case can be made for assignment and the
\k{assign} function. There is a dedicated \k{do.call} function to perform
function calls. However, the unnecessary uses of \eval are not easy to find and
often can only be uncovered by manual analysis. Dynamic analysis is limited due
to coverage issues, and static analysis of R remains an open problem.

@@

For the example in Listing~\ref{lst:unnecessary}, the function uses \eval to
lookup a target function name from a constructed string. Then it constructs a
call with the arguments that are passed to the current function. \Eval is
overkill in this case: a semantically equivalent function body can use \k{get}
to perform the lookup and then call the resulting function directly.

@@

\paragraph{Eval alternatives}

There are alternative implementations of \eval. The \k{lazyeval} package
provides \k{lazy} and \k{lazy_eval} as safer variants for the built-in
\k{substitute} and \k{eval} functions. The main differences are
that the \k{lazy} function captures not only the expression, but also the
environment associated with a function argument, and that it follows promises
across function invocations. This helps to avoid scoping bugs, as shown in the
example bellow:

@@

 x <- 2020
 f(year > x) # uses wrongly 1 for x
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.48\textwidth}
    \begin{lstlisting}[title={Use of \k{lazyeval}},label=lst:lazyeval]
 filter <- function(df, cond) {
   cond <- lazy(cond)
   r <- lazy_eval(cond, df)
   df[r, ]
 }

@@

 x <- 2020
 f(year > x) # uses correctly 2020 for x
\end{lstlisting}
\end{minipage}
\end{figure}

The \k{filter} function selects rows from \k{df} that match a given condition
passed in \k{cond}. However, the implementation using the built-in R \eval has a
scoping issue since it does not correctly capture the proper environment
associated with the condition.\footnote{This is a simplified version of the
  built-in \texttt{\scriptsize subset} function which suffers from this exact
problem.} The \k{lazyeval} version does not manifest the issue as it properly
captures the correct environment.

The tidyverse libraries\footnote{An opinionated collection of R packages
designed for data science. It consists of some of the most popular packages.}
include their own implementation of \eval called
\k{rlang::eval\_tidy}~\cite{tidyverse}, which departs from \eval in two ways.
First, it supports the evaluation of quosures, custom objects used for
metaprogramming that bundle expressions with an environment. The \k{eval\_tidy}
function accepts a data mask argument, a set of bindings such as a data frame
or list, which takes precedence over the environment. Effects such as
assignments happen in the data mask, and expressions such as \k{return()} do
not work since the data mask does not correspond to a frame on the call stack.

>>>>

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute}, and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use eval, we
manually inspected the top 117 sites, which contribute to 90\% of the calls in
the corpus. We present examples from nine classes from this batch. This
classification is not exhaustive since there are \PkgEvalCallSites \eval call
sites in our corpus, and many uses of eval do not fit in a specific category.
further categorize the nine classes into two groups: those that improve the
interface for users and those that simplify the implementation for developers.

\subsection{Better Interface}

This group contains \eval uses that enable the development of interfaces for
quick interactive data exploration.

@@

\begin{lstlisting}[title={\k{data.table::[}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
#    origin       V1
# 1:    LGA 11.35083
# 2:    EWR 14.96171
\end{lstlisting}\medskip

@@

\begin{lstlisting}[title={\k{glue::glue_data}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
\end{lstlisting}

@@

\paragraph{Extracting Varargs} \Eval is used to obtain the source code of the
arguments passed to a vararg parameter (\k{...}). In the example below,
\k{substitute(alist(...))} yields an expression representing a call to \k{alist}
with the \k{...} expanded to the arguments supplied to the function. Passing
this to \eval returns a list of the unevaluated argument expressions. This
pattern is needed to avoid premature evaluation of vararg arguments.

\begin{lstlisting}[title={\k{statnet.common::NVL}}, captionpos=b]
 function(...) {
   for (e in eval(substitute(alist(...)))) {
\end{lstlisting}

\paragraph{Controlled Evaluation} \Eval can be used to control when and if
expressions are evaluated. For example, \k{assertthat} allows users to make
assertions and, if they fail, produce legible error messages. The \texttt{\small
  assertthat::see\_if} function accepts a vararg with a list of expressions. It
extracts these expressions using the extracting varargs pattern and evaluates
them sequentially in the caller's environment using \k{eval}. If an expression
does not hold, an error is generated.

@@

\paragraph{Shadowing} First-class environment and \eval enable R users to
implement variations of core language functionality. For example, the \k{igraph}
package provides the \k{do_call} function shown below. This is a variant of the
built-in \k{do.call} function with different semantics for resolving the
function name. \k{do_call} calls a function (\k{f}) in the specified environment
(\k{env}) with the arguments supplied as a list (\k{args}).

@@

This group contains \eval uses that simplify implementation for package
developers.

\paragraph{Code Generation} \Eval can be used to generate code that would either
be tedious to write by hand or that is only known after deployment. The
following code snippet from the \k{Rcpp} package uses \eval to automatically
generate R bindings to C++ methods.

@@

\paragraph{Boilerplate Removal} A variant of code generation is to let users
write compact code that is expended via \eval and \k{match.call}. The example
uses \k{match.call} to access the expression with which the function is called.
The following three lines construct a call to \k{stats::model.frame} with
argument text of parameters \k{'formula'}, \k{'data'}, and \k{'weights'}. Then,
the call expression is evaluated in the parent environment. The use of
\k{match.call} avoids repeated use of \k{substitute} for extracting the argument
expressions. It is extensible as new parameter names can be easily added. This
example illustrates one of the most recurring uses of \eval by packages that do
statistical modeling. It is a typical example of the \k{model.frame} shape in
Table~\ref{tab:minimizedexpressions}.

@@

\paragraph{Obfuscation} A few packages use \eval to bypass the restrictions
imposed by the \k{R CMD CHECK} tool. This tool enforces some well-formedness
rules on a submitted package before accepting it for inclusion in CRAN. Some
static checks ensure that the package code does not use certain restricted
functions. Package authors use \eval to obfuscate their code to get around this
limitation. The example below mutates the variable \k{s} in the package
environment. However, it is locked by default, and unlocking it requires
\k{unlockBinding} which is a restricted function. To work around this, the call
to \k{unlockBinding} is done through \eval.

@@

In R, \eval is used chiefly for meta-programming and accessing remote
environments. Unlike JavaScript, which according to~\citet{ecoop11} had a small
set of well-defined patterns that could be rewritten without \eval, our results
suggest that \eval is an integral part of programming in R. However, there are
cases in which \eval is not needed, and other, more stylized functions can be
used. There are also alternatives to \eval that provide a more principled way
for meta-programming and dealing with lazy evaluation. In this section, we
briefly discuss these issues.

\paragraph{Unnecessary use of eval} The cases where \eval appears to be an
overkill include simple expression shapes such as \k{X}, \k{V}, \texttt{\$},
\k{<-}, and \k{FUN}. Variable lookup in a remote environment can be performed by
the built-in \k{get} function. A similar case can be made for assignment and the
\k{assign} function. There is a dedicated \k{do.call} function to perform
function calls. However, unnecessary uses of \eval are not easy to find and
often can only be uncovered by manual analysis. Dynamic analysis is limited due
to coverage issues, and static analysis of R remains an open problem.

@@

In the example in Listing~\ref{lst:unnecessary}, the function uses \eval to
lookup a target function name from a constructed string. Then it constructs a
call with the arguments that are passed to the current function. \Eval is
overkill in this case: a semantically equivalent function body can use \k{get}
to perform the lookup and then call the resulting function directly.

@@

\paragraph{Eval alternatives}
Alternative implementations of \eval are also available from R packages. The
\k{lazyeval} package provides \k{lazy} and \k{lazy_eval} as safer variants for
the built-in \k{substitute} and \k{eval} functions, respectively. These variants
differ from their counterparts in that they capture both the argument
expressions and their environments, and they follow promises across function
invocations. This helps to avoid scoping bugs, as shown in the example below.

@@

 x <- 2020
 f(year > x) # incorrectly uses 1 for x
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.48\textwidth}
    \begin{lstlisting}[title={Use of \k{lazyeval}},label=lst:lazyeval]
 filter <- function(df, cond) {
   cond <- lazy(cond)
   r <- lazy_eval(cond, df)
   df[r, ]
 }

@@

 x <- 2020
 f(year > x) # correctly uses 2020 for x
\end{lstlisting}
\end{minipage}
\end{figure}

The \k{filter} function selects rows from \k{df} that match a given condition
passed to \k{cond}. The implementation using the built-in \eval suffers from a
scoping bug since it always evaluates \k{cond} in the caller's environment
irrespective of where it is passed from.\footnote{This is a simplified version
  of the built-in \texttt{\scriptsize subset} function which suffers from this
  exact problem.} The \k{lazyeval} version avoids this problem since it
correctly captures the environment associated with \k{cond}.

The tidyverse libraries\footnote{An opinionated collection of R packages
  designed for data science. It consists of some of the most popular packages.}
include their own implementation of \eval called \k{eval\_tidy}~\cite{tidyverse}
from the \k{rlang} package. This function which departs from the built-in \eval
in two ways. First, it supports the evaluation of quosures, custom objects used
for metaprogramming that bundle expressions with an environment. Second, the
\k{eval\_tidy} function accepts a data mask argument, a set of bindings such as
a data frame or list, which takes precedence over the environment. Consequently,
effects such as assignments happen in the data mask, and expressions such as
\k{return()} do not work since the data mask does not correspond to a frame on
the call stack.
